Mortality Trend Among Shelter Residents: 

### 2024-09-27 paper/paper.qmd
final preread of the paper and edits

In this paper we will look at the data @toronto_deaths_shelter_residents, which explore the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates.The data indicates that mortality rates were relatively low, with fewer than five deaths per month in the prior 2019. However, after 2019, there was a sharp increase in the number of deaths, peaking in 2021 and 2022. This surge can be attributed to the onset of the COVID-19 pandemic in 2019, which significantly impacted public health and resulted in many fatalities.

@@

The data used in this paper is access in from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents). To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), tidyr (@tidyr), gridExtra (@gridExtra), knitr (@knitr), and ggplot2 (@ggplot2). LLM (@openai2023chatgpt) was also use for this paper in term of helping with coding in R to graph data, simulating a data, cleaning data, and revising writing.

@@

The Deaths of Shelter Residents data is part of a larger database shared by Toronto Public Health (TPH), which also includes deaths of homeless individuals living in shelters and outdoors (@toronto_homeless_deaths). Therefore, the findings in this paper focus solely on homeless individuals residing in shelters, which may limit the data set and lead to potential biases that do not fully represent the mortality of all homeless individuals.

@@

The following graph (@fig-TotalNumberofDeathByYear) shows the overall trend in the number of decedents living in shelters from 2007 to 2024. It reveals that between 2007 and 2013, the number of deaths remained quite low, under 30 each year. However, starting in 2013, the number of deaths increased, and the mortality rate never returned to the low levels seen from 2007 to 2013. The graph also highlights two significant peaks in 2015 and 2021. The increase in deaths was more pronounced in 2021 compared to 2015. Further discussion on the reasons for these peaks can be found in discussion sections (@sec-first-dpoint and @sec-second-dpoint). Furthermore, in the appendix (@sec-app), there are graph for number of death per month for each individual year.

@@

# Create the bar graph with adjusted year labels
plot_data |>
  ggplot(aes(x = year_short, y = Mortality, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  labs(
    title = "Mortality Numbers of Male Vs Female Decedents over 2007 to 2024",
    x = "Year (2000s)",
    y = "Number of Decedents"
  ) +
  scale_fill_manual(values = c("Male_decedents" = "blue", "Female_decedents" = "pink")) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```
@fig-AvgAgeofDecedentsFemaleVSMale further examines the differences in average age at death between females and males. It's important to note that there are significantly more male deaths than female deaths, which could result in a lower average age for males due to the larger sample size. Although the graph does not definitively indicate whether males generally die at an older age than females or vice versa, it does suggest that the average age at death for males appears to be considerably higher than that for females during many periods indicated in the graph. More information is available in @sec-3-dpoint.

@@

>>>>

In this paper we will look at a data (@toronto_deaths_shelter_residents), which explore the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates. The data indicates that mortality rates were relatively low, with fewer than five deaths per month prior of 2019. However, after 2019, there was a sharp increase in the number of deaths, peaking in 2021 and 2022. This surge can be attributed to the onset of the COVID-19 pandemic in 2019, which significantly impacted public health and resulted in many fatalities.

@@

The data used in this paper is access from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents). To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), tidyr (@tidyr), gridExtra (@gridExtra), knitr (@knitr), and ggplot2 (@ggplot2). LLM (@openai2023chatgpt) was also use in term of helping with coding in R to graph data, simulating data, cleaning data, and revising writing.

@@

The Deaths of Shelter Residents data is part of a larger database shared by Toronto Public Health (TPH), which includes deaths of homeless individuals living in shelters and outdoors (@toronto_homeless_deaths). Therefore, the findings in this paper focus solely on homeless individuals residing in shelters, which limit the data set and lead to potential biases that do not fully represent the mortality of all homeless individuals.

@@

The following graph (@fig-TotalNumberofDeathByYear) shows the overall trend in the number of decedents living in shelters from 2007 to 2024. It reveals that between 2007 and 2013, the number of deaths remained quite low, under 30 death each year. However, starting in 2013, the number of deaths increased, and the mortality rate never returned to the low levels seen from 2007 to 2013. The graph also highlights two significant peaks in 2015 and 2021. The increase in deaths was more pronounced in 2021 compared to 2015. Further discussion on the reasons for these peaks can be found in discussion sections (@sec-first-dpoint and @sec-second-dpoint). Furthermore, in the appendix (@sec-app), there are graph for number of death per month for each individual year.

@@

# Create the bar graph with adjusted year labels
plot_data |>
  ggplot(aes(x = year_short, y = Mortality, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  labs(
    title = "Mortality Numbers of Male Vs Female Decedents over 2007 to 2024",
    x = "Year (2000s)",
    y = "Number of Decedents"
  ) +
  scale_fill_manual(values = c("Male_decedents" = "blue", "Female_decedents" = "pink")) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```
@fig-AvgAgeofDecedentsFemaleVSMale further examines the differences in average age at death between females and males. It is important to note that there are significantly more male deaths than female deaths, which could result in a lower average age for males due to the larger sample size. Although the graph does not definitively indicate whether males generally die at an older age than females or vice versa, it does suggest that the average age at death for males appears to be considerably higher than that for females during many periods indicated in the graph. More information is available in @sec-3-dpoint.

@@

In summary, the paper does not provide statistical evidence for all the possible reasons discussed regarding the mortality trends of homeless individuals residing in shelters. However, it can serve as a guide and signal for city organizations to enhance their efforts in supporting the homeless population.
